---
title: "working_to_final"
output: word_document
date: "2025-10-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(out.width = "60%",fig.align='center')
```

1. Ethan + Sydney's data cleaning and transformation conbined here.
## Load Libraries 
```{r}
# install.packages("here")   # only once allows us to all call from same space 
# install.packages("ResourceSelection")
# install.packages("randtests")
```
```{r}
library(tidyverse)
library(knitr)
library(tinytex) 
library(dplyr)
library(ggplot2)
library(readr)
library(tibble)
library(here)
library(ResourceSelection)
library(randtests)
library(purrr)
library(stringr)
library(mgcv)      # for GAM
library(glmmTMB)   # for mixed-effects logistic regression
```

## Load Datasets
```{r}
volley_points <- readr::read_csv(here("datasets", "volleyball_points_dataset.csv"))  
all_matches_CU <- readr::read_csv(here("datasets", "all_matches_CU.csv"))                #551 by 16
all_matches_opp <- readr::read_csv(here("datasets","all_matches_opp.csv"))               #551 by 16
all_matches_processed <- readr::read_csv(here("datasets","all_matches_processed.csv"))  #1102 by 16
all_matches <- readr::read_csv(here("datasets","all_matches (1).csv"))                 #269628 by 125 
```

##DATA CLEANING 
```{r}
#data inflow + outflow
in_path <- here::here("datasets", "all_matches (1).csv")
out_path <- "volleyball_points_dataset.csv"
```

```{r}
touches <- readr::read_csv(in_path, show_col_types = FALSE)

#normalize season column
touches <- touches %>%
  mutate(season = str_extract(season, "\\d{2}$"))

order_col <- if ("Unnamed: 0" %in% names(touches)) "Unnamed: 0" else NULL

#touches in correspondence to rally order   
touches_ord <- touches %>%
  {
    if (!is.null(order_col)) arrange(., match_id, set_number, .data[[order_col]])
    else group_by(., match_id, set_number) %>% arrange(row_number(), .by_group = TRUE) %>% ungroup()
  }
```
To compress rallies accurately, touches must be sorted in the exact sequence they occurred. We sort by match_id, set_number, and a stable row index so subsequent steps (like detecting score changes) behave correctly.

```{r}
#add flags for rally IDS + rally-ending scores 
touches_flagged <- touches_ord %>%
  filter(!is.na(home_team_score) | !is.na(visiting_team_score))  %>% # remove empty rows early 
  group_by(match_id, set_number) %>%
  mutate(
    prev_home  = lag(home_team_score),
    prev_away  = lag(visiting_team_score),
    rally_end_flag = (home_team_score != prev_home) | (visiting_team_score != prev_away),
    rally_end_flag = replace_na(rally_end_flag, TRUE),
    rally_id = cumsum(rally_end_flag),
    match_end_flag = last(to_logical_vec(won_match))  # NEW: indicates if this match was won by the "home" team 
    #TO LOOK AT WHOLE MATCH WIN PROB NOT JUST RALLY/SETS
  ) %>%
  ungroup()
```
Volleyball is rally scoring; each rally ends when the scoreboard moves. By turning score changes into a “flag” and cumulatively summing, every touch is assigned to the rally that produced (or followed) a point.

```{r}
#function to create binaries (T/F)
to_logical_vec <- function(x) {
  if (is.logical(x)) x
  else if (is.numeric(x)) x != 0
  else tolower(as.character(x)) %in% c("true","t","yes","y","1")
}
```
“Between points” events like timeouts are often recorded right after a point and before the next serve. We’ll roll them up with the rally that just ended (i.e., the point that triggered them).

```{r}
#collapse touches into rally-level dataset
points_core <- touches_flagged %>%
  group_by(match_id, set_number, rally_id) %>%
  summarise(
    home_score_end           = last(home_team_score),
    away_score_end           = last(visiting_team_score),
    serving_team             = first(serving_team),          
    touches_in_rally         = n(),
    team_rotation            = suppressWarnings(as.integer(last(team_rotation))),
    opp_rotation             = suppressWarnings(as.integer(last(opp_rotation))),
    score_diff_raw           = suppressWarnings(as.integer(last(score_diff))),
    timeout_after_point      = any(to_logical_vec(timeout), na.rm = TRUE),
    substitution_after_point = any(to_logical_vec(substitution), na.rm = TRUE),
    #home_team_won_rally      = last(to_logical_vec(home_team_won)),  # new
    won_match_rally          = last(to_logical_vec(won_match)),      # new
    home_team                = last(home_team),
    visiting_team            = last(visiting_team),
    won_set                  = last(won_set),
    date                     = last(date),
    season                   = last(season),
    conference               = last(conference),
    .groups = "drop"
  ) %>%
  arrange(match_id, set_number, rally_id)
```
(JUST NOTED: Touches in rally is incorrect due to nature of all-matches data set with "empty" rows)
-We collapse many touches into a single rally row. We take last scores to capture the scoreboard after the point, and any() over timeout/sub to mark whether the inter-point stoppage happened following that point.

```{r}
#add point winner info. 
points_w_winner <- points_core %>%
  group_by(match_id, set_number) %>%
  mutate(
    prev_home_end = lag(home_score_end, default = 0),
    prev_away_end = lag(away_score_end, default = 0),
    home_delta    = home_score_end - prev_home_end,
    away_delta    = away_score_end - prev_away_end,
    point_winner_team = case_when(
      home_delta == 1 & away_delta == 0 ~ home_team,
      away_delta == 1 & home_delta == 0 ~ visiting_team,
      TRUE ~ NA_character_            
    )
  ) %>%
  ungroup()
```
Had some problems with timeout booleans. So above is the fix. Exactly one side’s score should increase by 1 at a rally end. Using deltas bypasses missing point_won_by values and avoids problems assigning winners when TO/Sub rows appear.

```{r}
#CU focused variables 
points_cu <- points_w_winner %>%
  mutate(
    cu_is_home        = home_team == "CU",
    cu_score_end      = if_else(cu_is_home, home_score_end, away_score_end),
    opp_score_end     = if_else(cu_is_home, away_score_end, home_score_end),
    cu_point_won      = point_winner_team == "CU",
    cu_score_diff_end = cu_score_end - opp_score_end,
    cu_serving        = serving_team == "CU"
  )
```
By aligning columns to CU, we can analyze run lengths, in-rally momentum, and timeout efficacy without worrying about home/away flipping your signs.

**NEW SECTIONS
```{r}
#pre-rally state variables (for transitions)
points_cu <- points_cu %>%
  group_by(match_id, set_number) %>%
  arrange(rally_id) %>%
  mutate(
    cu_score_pre  = lag(cu_score_end, default = 0),
    opp_score_pre = lag(opp_score_end, default = 0),
    d_pre         = cu_score_pre - opp_score_pre
  ) %>%
  ungroup()
```

```{r}
#set/match + deciding vs nondeciding set 
points_cu <- points_cu %>%
  group_by(match_id) %>%
  mutate(
    total_sets = max(set_number, na.rm = TRUE),
    deciding_set = (set_number == total_sets),
    rally_in_match = row_number(),
    rally_prop_in_match = rally_in_match / n()
  ) %>%
  ungroup()
```

```{r}
#high leverage points 
points_cu <- points_cu %>%
  mutate(
    rally_importance = case_when(
      cu_score_end >= 20 | opp_score_end >= 20 ~ "late_set",
      TRUE ~ "early_mid_set"
    ),
    # Flag rallies where the score difference is small (≤ 2 points)
    #“close”//high-pressure moments
    close_score = abs(cu_score_diff_end) <= 2
  )
```

*may not use below but good to have.
```{r}
#momentum/clustering variables (detect run-length effects)
points_cu <- points_cu %>%
  group_by(match_id, set_number) %>%
  arrange(rally_id) %>%
  mutate(
    prev_win = lag(cu_point_won),
    # Assign a unique run ID whenever the rally result switches (win → loss)
    run_id = cumsum(cu_point_won != lag(cu_point_won, default = cu_point_won[1])),
    #Count how many rallies are in each consecutive run
    run_length = ave(seq_along(run_id), run_id, FUN = length)
  ) %>%
  ungroup()
```
** NEW SECTIONS DONE.

```{r}
#final set-level scores 
set_scores <- touches_flagged %>%
  group_by(match_id, set_number) %>%
  summarise(
    final_home_set_score = max(home_team_score, na.rm = TRUE),
    final_away_set_score = max(visiting_team_score, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
points_final <- points_cu %>%
  left_join(set_scores, by = c("match_id", "set_number")) %>%
  select(
    match_id, date, season, conference,
    home_team, visiting_team, set_number, rally_id,
    cu_serving,
    cu_score_pre, opp_score_pre, d_pre,   # NEW
    cu_score_end, opp_score_end, cu_score_diff_end, cu_point_won,
    deciding_set, rally_importance, close_score, rally_in_match, rally_prop_in_match, run_length, prev_win, # NEW CONTEXT
    final_home_set_score, final_away_set_score,
    won_set,
    touches_in_rally, timeout_after_point, substitution_after_point,
    won_match_rally
  )
```

Final set scores provide outcome context for every rally (like late-set pressure).
```{r}
#output dataset
write_csv(points_final, out_path)

points <- points_final %>%
  na.omit()
```

```{r}
#export to github
readr::write_csv(points_final, "points_final.csv")
readr::write_csv(points_final, "points.csv")
```

Points_final: 27580, 30 variables 
variables: match_id, date, season, conference, home_team, visiting_team, set_number, rally_ud, cu_serving, cu_score_pre, opp_score_pre, d_pre, cu_score_end, opp_score_end, cu_score_diff_end, cu_point_won, deciding_set, rally_importance, close_score, rally_in_match, rall_prop_in_match, run_length, prev_win, final_home_set_score, final_away_set_score, won_set, touches_in_rally, timeout_after_point, substitution_after_point, won_match_rally

Points: 23365 entries , 30 variables (NO NAs)
variables: match_id, date, season, conference, home_team, visiting_team, set_number, rally_ud, cu_serving, cu_score_pre, opp_score_pre, d_pre, cu_score_end, opp_score_end, cu_score_diff_end, cu_point_won, deciding_set, rally_importance, close_score, rally_in_match, rall_prop_in_match, run_length, prev_win, final_home_set_score, final_away_set_score, won_set, touches_in_rally, timeout_after_point, substitution_after_point, won_match_rally

*I THINK WE SHOULD USE POINTS AS OUR DATASET FOR ANY FUTURE WORK*.

##FUTURE WORK ADDED 






